---
title: "Data Cleaning"
format: html
---

# Packages
```{r}
library(tidyverse)
```


# Read CSV
```{r}
cry_df <- read_csv("./Data/full_data_mod.csv")
head(cry_df)
```

# Clean Data

Find the NAs/missing values
```{r}
cry_df %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "column", values_to = "na_count") %>%
  arrange(desc(na_count))

cry_df %>%
  group_by(ID) %>%
  summarise(count = n()) %>%
  arrange(desc(count))

cry_df %>%
  group_by(Reason) %>%
  summarise(count = n()) %>%
  arrange(desc(count))

cry_df %>%
  group_by(Age) %>%
  summarise(count = n()) %>%
  arrange(desc(count))

cry_df %>%
  group_by(Gender) %>%
  summarise(count = n()) %>%
  arrange(desc(count))
```

We can see that we haave 260 missing emails, 380 missing age, and 313 missing gender. That's around 1/4 of the data. We also have this thing called MULT-REDCAP-GENDER which is confusing.

Cleanup the Sample Column: 
```{r}
cry_df <- cry_df %>%
   mutate(Sample = str_remove(Sample, "\\.wav$"))
```

Numerical Transformations: Make the Numerical Values all numeric,
```{r}
cry_df <- cry_df %>%
  mutate(Age = ifelse(Age == "NO-AGE", NA, Age), 
         Gender = ifelse(Gender== "NO-GENDER", NA, Gender)) %>% # turn into NA
  mutate(Age = as.numeric(Age), Sample = as.numeric(Sample))
```

Categorical Transformations: Make a new ID label, cleanup parentfile, turn categorical variables into factors
```{r}
cry_df <- cry_df %>%
  mutate(newID = row_number(),
         Reason = as.factor(Reason),
         Gender = as.factor(Gender),
         ParentFile = as.factor(str_match(ParentFile, "_([^_]+)\\.csv")[,2])) %>%
  relocate(newID, .before = ID)
```

```{r}
glimpse(cry_df)
```

Make a different dataframe with NA's dropped
```{r}
cry_df_na_drop <- cry_df %>%
  drop_na()
```

Write to a CSV
```{r}
write_csv(cry_df, "full_data_cleaned_include_nas.csv")
write_csv(cry_df_na_drop, "full_data_cleaned_without_nas.csv")
```


