---
title: "Survey Design Review Exercise"
format: pdf
---

```{r}
library(psych)
library(paran) 
library(MASS)
library(GPArotation)
library(tidyverse)
library(reshape2)
library(ggplot2)
```

## Running Cronbach alpha 

```{r}
stemq2 <- read_csv("./stemq2.csv")

# Inspect to confirm it worked
glimpse(stemq2)
```


```{r}
psych::alpha(stemq2)
```


## Finding the correlation between each item and the total score

```{r}
attach(stemq2)
totalscore <- (q1+q2+q3+q4+q5+q6+q7+q8+q9+q10+q11+q12+q13+q14+q15+q16+q17+q18+q19+q20+q21+q22+q23+q24)
newdata <- data.frame(q1,q2,q3,q4,q5,q6,q7,q8,q9,q10,q11,q12,q13,q14,q15,q16,q17,q18,q19,q20,q21,q22,q23,q24,totalscore)
detach(stemq2)
head(newdata)

cor_matrix <- cor(newdata)
cor_totalscore <- cor_matrix["totalscore", , drop=FALSE]
cor_totalscore

cor_df <- as.data.frame(cor_totalscore) %>%
  pivot_longer(cols = everything(), names_to = "Question", values_to = "Correlation with Totalscore") %>%
  arrange(`Correlation with Totalscore`) %>%
  mutate(Question = factor(Question, levels = Question))  # enforce ordering

ggplot(cor_df, aes(x = `Correlation with Totalscore`, y = Question)) +
  geom_col(fill = "steelblue") +
  xlim(0, 1) +
  theme_minimal() +
  labs(title = "Correlations with Totalscore", x = "Correlation", y = "Question")
```
 
## Running parallel analysis

```{r}
paran(stemq2,cfa=TRUE, graph=TRUE,color=TRUE,col=c("black","red","blue"))
```


## R command for running exploratory factor analysis

```{r}
fa(stemq2,nfactors=2,rotate="oblimin")
```

## command for running Kaiser-Meyer-Olkin factor adequacy - a value KMO values between 0.8 and 1indicate the sampling is adequate. KMO values less than 0.6 indicate the sampling is not adequate - some set the cut off at 0.50. We get a value of 0.82, so we are OK

```{r}
KMO(stemq2)
```

## Command for creating explanatory factor analysis (let us assume three underlying factors). Factors one, two, and three explain 60%, 24%, and 16% of the variance respectivley.

```{r}
fa(stemq2,nfactors = 2,rotate="oblimin")
```


## Command for creation of diagram underlying extracted factors 

```{r}
M1<-fa(stemq2,nfactors = 2,rotate="oblimin")
fa.diagram(M1,main="stemq2")
```


